Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-25 paper/paper.tex
functorial actions side by side
some work on the bib
small tweaks

Writing a parser as a linear term is an intrinsic
verification of the \emph{soundness} of the parser completely for free
from the typing: any $\inl$ parse that we return \emph{must}
correspond to a parse tree of the input string. Further, if we verify
the disjointness property of \cref{def:parser} we then also get the \emph{completeness} of
the parser as well, that it never fails to generate an $A$ parse when
it is possible.

@@

\section{Discussion and Future Work}
\label{sec:discussion}
\paragraph{Grammars as (Linear) Types}
Lambek's original syntactic calculus \cite{lambek58} describes a
logical system for linguistic derivations, and it can be given
semantics inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. This led to many uses of non-commutative
linear logic and lambda calculi in linguistics
\cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style
of grammar formalism has gone by the names Lambek calculus or
\emph{categorial grammar}, and it is equal in expressivity to
context-free grammars.  The existing works on categorial grammar are
different in nature to our approach: they are based on non-commutative
linear logic, but their terms do not include elimination rules, and so
can only express parse trees, and so cannot be used to express
verified parsers.

@@

%% A frequently fruitful research direction is exploring varying
%% extensions of Kleene algebras, Kleene algebra with tests (KAT) being
%% one of the most notable ones \cite{kozen1997kleene}. Our approach is
%% radically different from most extensions, which usually aim at
%% modifying or adding new operations to Kleene algebras, but still
%% keeping it as an inequational theory. By adopting a category-theoretic
%% treatment and allowing the ``order structure'' to encode more
%% information than merely inequalities, we were able to extend Kleene
%% algebra to reason about parsing as well.
\paragraph{Dependent Linear Types}
Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
distinct feature of these systems is that linear types and non-linear types are
distinct sorts, so the linear logic $!$ operator is not a primitive, but is
instead definable, for instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.
V\'ak\'ar develops a dependent linear type theory where instead the non-linear
types are accessed using the $!$ modality, similar to Girard's original approach
\cite{vakar,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
categorical semantics for their system, whereas we have only developed a single
denotational model. It appears straightforward to adapt V\'ak\'ar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.

@@

\ifappendix
\appendix
\section{Syntax}
\begin{figure}
  \begin{align*}
    \el(\Var\,M) B &= B\,M\\
    \el(\mathsf{K}\,A) B &= A\\
    \el(\bigoplus A) B &= \bigoplus_{y:Y}\el(A y)B\\
    \el(\bigamp A) B &= \bigamp_{y:Y}\el(A y)B\\
    \el(A \otimes A') B &= \el(A)B \otimes \el(A')B\\\\
    \map(\Var\,M)\,f &= f\,M\\
    \map(\K\,A)\,f &= \lambda a. a\\
    \map(\bigoplus A)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
    \map(\bigwith\,A)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
    \map(A \otimes A')\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
  \end{align*}
  \caption{Strictly positive functors functorial actions}
\end{figure}

@@

\begin{figure}
  \begin{mathpar}
    \boxed{
      \inferrule
      {\linterm \Gamma \Delta e A \and
       \linterm \Gamma \Delta e' A}
      {\Gamma ; \Delta \vdash e \equiv e' : A}}

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{\Gamma ; \Delta , a : A \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta,\Delta' \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta \vdash e : A \lto B}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \apptol {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\apptol e a} : B \tol A}
%
    \and
%
    \inferrule{\Gamma , x : X \vdash e : A \\ \Gamma \vdash M : X}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
% %
%     \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
% %
%     \and
% %
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{\Gamma ; \Delta_1 , \Delta_2 \vdash e : C}{\Gamma; \Delta_1 , \Delta_2 \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1 , a : A , \Delta_3 \vdash e' : C}{\Gamma; \Delta_1 , \Delta_3 \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \\ \Gamma ; \Delta_3 \vdash e' : B \\ \Gamma ; \Delta_1 , a : A , b : B , \Delta_4 \vdash e'' : C}{\Gamma; \Delta_1 , \Delta_2, \Delta_3 , \Delta_4 \vdash \letin {(a, b)} {(e , e')} e'' \equiv e'' \{ e/a , e'/b \} : C}
%
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1 , c : A \otimes B , \Delta_3 \vdash e' : C }{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {\subst {e'} {(a, b)} c} \equiv \subst {e'} e c : C}
%
    \and
%
    \inferrule{\Gamma \vdash M : X \\ \Gamma ; \Delta_2 \vdash e : A \\ \Gamma , x : X \vdash \Delta_1 , a : A , \Delta_3 }{\Gamma;\Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} {\sigma~M~e} {e'} \equiv e' \{ M/x , e/a \} : C}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_1, y : \LinSigTy {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTy {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta \vdash f~e \equiv g~e}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer e f g}
    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e :
      \equalizer e f g}
\end{mathpar}
  \caption{Judgmental equality for linear terms}
  \label{fig:jdgeq}
\end{figure}

>>>>

Writing a parser as a linear term intrinsically verifies the \emph{soundness} of
the parser for free from the typing: any $\inl$ parse that we return \emph{must}
correspond to a parse tree of the input string. Further, if we verify the
disjointness property of \cref{def:parser} we then also get the
\emph{completeness} of the parser as well, that when the parser rejects the
input that there are no valid parses.

@@

\section{Related and Future Work}
\label{sec:discussion}

\subsection{Related Work}
\paragraph{Grammars as (Linear) Types}
Lambek's original syntactic calculus \cite{lambek58} describes a
logical system for linguistic derivations, and it can be given
semantics inside of a non-commutative biclosed monoidal category
\cite{lambek1988categorial}. This led to many uses of non-commutative
linear logic and lambda calculi in linguistics
\cite{buszkowskiTypeLogicsGrammar2003} --- including mechanized
categorial grammar parsers \cite{Guillaume2024,ranta-2011}. This style
of grammar formalism has gone by the names Lambek calculus or
\emph{categorial grammar}, and it is equal in expressivity to
context-free grammars.  The existing works on categorial grammar are
different in nature to our approach: they are based on non-commutative
linear logic, but their terms do not include elimination rules, and so
can only express parse trees, and so cannot be used to express
verified parsers.

@@

%% A frequently fruitful research direction is exploring varying
%% extensions of Kleene algebras, Kleene algebra with tests (KAT) being
%% one of the most notable ones \cite{kozen1997kleene}. Our approach is
%% radically different from most extensions, which usually aim at
%% modifying or adding new operations to Kleene algebras, but still
%% keeping it as an inequational theory. By adopting a category-theoretic
%% treatment and allowing the ``order structure'' to encode more
%% information than merely inequalities, we were able to extend Kleene
%% algebra to reason about parsing as well.
\paragraph{Dependent Linear Types}
Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
Wadler's linear-non-linear calculus \cite{benton1994}. A
distinct feature of these systems is that linear types and non-linear types are
distinct sorts, so the linear logic $!$ operator is not a primitive, but is
instead definable, for instance in \theoryabbv as $!A = \LinSigTy{\_}{\ltonl A}{I}$.
V\'ak\'ar develops a dependent linear type theory where instead the non-linear
types are accessed using the $!$ modality, similar to Girard's original approach
\cite{vakar2015,girard_linear_1987}. Additionally, V\'ak\'ar develops a general
categorical semantics for their system, whereas we have only developed a single
denotational model. It appears straightforward to adapt V\'ak\'ar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.

@@

\ifappendix
\appendix
\section{Syntax}
\begin{figure}
\begin{minipage}[t]{.35\textwidth}
  \begin{align*}
    \el\left(\Var\,M\right) B &= B\,M\\
    \el\left(\mathsf{K}\,A\right) B &= A\\
    \el\left(\bigoplus A\right) B &= \bigoplus_{y:Y}\el(A y)B\\
    \el\left(\bigamp A\right) B &= \bigamp_{y:Y}\el(A y)B\\
    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B
  \end{align*}
\end{minipage}%
\begin{minipage}[t]{.65\textwidth}
\begin{align*}
    \map\left(\Var\,M\right)\,f &= f\,M\\
    \map\left(\K\,A\right)\,f &= \lambda a. a\\
    \map\left(\bigoplus A\right)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
    \map\left(\bigwith\,A\right)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
    \map\left(A \otimes A'\right)\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
  \end{align*}
\end{minipage}%
  \caption{Strictly positive functors functorial actions}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \boxed{
      \inferrule
      {\linterm \Gamma \Delta e A \and
       \linterm \Gamma \Delta e' A}
      {\Gamma ; \Delta \vdash e \equiv e' : A}}

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{\Gamma ; \Delta , a : A \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta,\Delta' \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta \vdash e : A \lto B}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{\Gamma ; a : A , \Delta \vdash e : C \\ \Gamma ; \Delta' \vdash e' : A }{\Gamma; \Delta , \Delta' \vdash \apptol {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A}{\Gamma; \Delta \vdash e \equiv \lambtol a {\apptol e a} : B \tol A}
%
    \and
%
    \inferrule{\Gamma , x : X \vdash e : A \\ \Gamma \vdash M : X}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTyLimit {x} {X} {A}}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTyLimit x X A}
%
    \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
% %
%     \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
% %
%     \and
% %
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{\Gamma ; \Delta_1 , \Delta_2 \vdash e : C}{\Gamma; \Delta_1 , \Delta_2 \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1 , a : A , \Delta_3 \vdash e' : C}{\Gamma; \Delta_1 , \Delta_3 \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
%
    \and
%
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \\ \Gamma ; \Delta_3 \vdash e' : B \\ \Gamma ; \Delta_1 , a : A , b : B , \Delta_4 \vdash e'' : C}{\Gamma; \Delta_1 , \Delta_2, \Delta_3 , \Delta_4 \vdash \letin {(a, b)} {(e , e')} e'' \equiv e'' \{ e/a , e'/b \} : C}
%
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1 , c : A \otimes B , \Delta_3 \vdash e' : C }{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {\subst {e'} {(a, b)} c} \equiv \subst {e'} e c : C}
%
    \and
%
    \inferrule{\Gamma \vdash M : X \\ \Gamma ; \Delta_2 \vdash e : A \\ \Gamma , x : X \vdash \Delta_1 , a : A , \Delta_3 }{\Gamma;\Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} {\sigma~M~e} {e'} \equiv e' \{ M/x , e/a \} : C}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_1, y : \LinSigTyLimit {x} {X} {A}, \Delta_2 \vdash e' : C \\ \Gamma ; \Delta_2 \vdash e : \LinSigTyLimit {x} {X} {A}}{\Gamma; \Delta_1 , \Delta_2 , \Delta_3 \vdash \letin {\sigma~x~a} e {\subst {e'} {\sigma~x~a} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta \vdash f~e \equiv g~e}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer e f g}
    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e :
      \equalizer e f g}
\end{mathpar}
  \caption{Judgmental equality for linear terms}
  \label{fig:jdgeq}
\end{figure}
